---
title: "Comparing Random Forest Approaches to Class-Imbalanced Credit Card Fraud Detection"
author: "Nicholas Pargoff and Andrew Mattson (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

Payment fraud is becoming more prevalent with the current climate of
e-commerce and the development of different payment methods. In July
2023, the Federal Reserve launched the FedNow Service, which allows
banks and credit unions of any size to send instant payments at any time
of day and every day of the year [@frbservices2023fednow]. As
real-time payment adoption increases, fraud detection models must flag
suspicious transactions as they happen. This increases the need for more
robust ways to identify transaction fraud. Random forest is commonly
used to address this problem. Current research focuses on handling the
unbalanced nature of fraudulent transactions. In this paper we test some
of the common random forest models against the original algorithm to
find which model has greater fraud detection performance.

Literature Review                                                                        

Random Forest (RF) is a widely adopted machine learning method because
of its computational efficiency and versatility [@scornet2026theory;
@salman2024random]. It uses random samples of data to build multiple
decision trees, and because the trees are not highly correlated the risk
of overfitting is relatively low [@salman2024random]. RF was initially
created to run classification and regression tests but has expanded into
other domains like survival analysis and quantile regression. The
subsampling structure allows the uncertainty of its predictions to be
quantified [@scornet2026theory]. RF’s main limitation is prediction
speed. It can train quickly but can be slow when needed for real-time
decisions [@salman2024random].

RF has proven effective for credit card fraud detection [@jebaseeli2020fraud; @borketey2024; @salunke2025fraud]. @borketey2024
compared it with logistic regression, linear discriminant analysis,
k-nearest neighbors, and classification trees on a dataset with 0.17%
fraudulent transactions with RF detecting about 71% of those
transactions. Some studies used RF that did not adjust for class
imbalance and the authors judged RF mainly on overall predictions
[@jebaseeli2020fraud]. RF has limitations like degrading when the
scale of the dataset increases [@xuan2018random]. Fraud changes over
time and data collection is different by region making it difficult to
generalize [@borketey2024].

Using accuracy as a metric is not the best way to measure performance of
this method when fraudulent transactions are rare. A model can have a
very high accuracy score while missing most of the fraudulent
transactions. For this reason, several studies relied on precision,
recall, F1, and area-under-the-curve measures [@galar2012review;
@borketey2024; @wijaya2024comparative; @aburbeian2023credit].

A common way to deal with data imbalance is to pair RF with SMOTE
[@bhagat2015enhanced; @aburbeian2023credit; @zhang2024improved]. This
combination has improved the minority-class detection outside fraud.
@bhagat2015enhanced applied SMOTE after splitting the multi-class
problem with One-vs-All and One-vs-One methods and @zhang2024improved combined
SMOTE with tree pruning and Z-score normalization and reached 98.03%
accuracy and 100% recall. The results do support the general approach of
pairing SMOTE with RF; however, this may not directly transfer over to
fraud detection. SMOTE was found to not always be the most effective
option. Random oversampling with XGBoost achieved the highest F1-score,
92.43% and the authors concluded that random oversampling, or
undersampling with NCR, gave the best results [@wijaya2024comparative].

Some methods change how the ensemble samples the data. @galar2012review reviewed 22 ensemble variants across four families, using a
single C4.5 decision tree as the base classifier with AUC for
evaluation. The most complex approaches did not always have the best
performance. SMOTEBagging, RUSBoost, and UnderBagging were the most
robust. RUSBoost provided good performance and was the least complex. @gu2022novel proposed that an improper sampling strategy can negatively
impact model performance, increase complexity, and cause overfitting.
The study also found that DCI-ISSA-RF, which combines DCI oversampling
with ISSA hyperparameter optimization, outperformed undersampling, most
cost-sensitive methods and baseline RF. However, DCI-ISSA-RF ran longer
on large credit card datasets which relates to the speed concern raised
by @salman2024random.

These studies find that RF is an acceptable choice for fraud detection.
They also provide different techniques for handling class imbalance. The
results are difficult to compare because the studies had significant
variations like different datasets and performance metrics. Some of the
studies did not use cross validation [@wijaya2024comparative] while others
used overall accuracy [@salunke2025fraud] for their results making it
unclear how each imbalance technique improves RF itself. This study
attempts to address the previous study variations by using a baseline RF
and only changing the imbalance treatment. Each approach will be
evaluated on fraud-class precision, recall, F1, and AUC-PR.

## Methods

- Detail the models or algorithms used.

- Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}$$ $W_n(x)$ *is the sum
of weights that belongs to all real numbers. Weights are positive
numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i$$

## Analysis and Results

### Data Exploration and Visualization

- Describe your data sources and collection process.

- Present initial findings and insights through visualizations.

- Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

- Explain your data preprocessing and cleaning steps.

- Present your key findings in a clear and concise manner.

- Use visuals to support your claims.

- **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

- Summarize your key findings.

- Discuss the implications of your results.

## References
