---
title: "Comparing Random Forest Approaches to Class-Imbalanced Credit Card Fraud Detection "
subtitle: "This is a Report Template Quarto"
author: "Nicholas Pargoff and Andrew Mattson (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report! I am adding this to show updates from RStudio.
:::

## Introduction

Card fraud is becoming more prevalent with the current climate or
e-commerce and the shifting of the US banking legacy systems over to
more cloud based real time transactions. This increases the need for a
more robust way to identify transaction fraud. Randon forest has been
seen as a way to try to combat this current issue. Research lies in
trying to find a way to deal with the unbalanced nature of fraudulent
transactions. Less than 2 percent of all transactions are fraudulent so
this requires a model that can handle unbalanced data in real time. Most
research looks at creating models that can handle this. This paper will
look at the common methods and validate their performance against a
nonadjusted random forest model. In this paper we are going to test some
of the common random forest models against the original algorithm to
test if these different models have a greater accuracy impact.

## Literature Review

In Random Forest Algorithm Overview, Salman et al. (2024) provide an
in-depth review of the benefits and uses of the random forest model,
which is a predictive machine learning model. The basis of the model is
that it creates decision trees from subsets of data. Having more trees
reduces variance. Important features of RF are that they are highly
accurate, are rarely overfitted, and can handle data with missing
values. Although RF has a lot of advantages, a major constraint of the
tool is its inability to produce results efficiently in real-time due to
reduced prediction speed after training.

*Theory of Random Forests* indicates that RF is the most widely adopted
machine learning tool due to its versatility for handling various
problems as well as computational efficiency. The mathematical theory of
RF has lagged even though the tool is highly popular and widely used
(Scornet & Hooker, 2026). Random forests were initially created to run
classification and regression tests but have expanded into other
domains. By altering how the data is reflected in the model, researchers
have brought RF into other domains like survival analysis and quantile
regression. Research on quantifying RF output has also increased. The
central limit theorem and variance of RF are products of the forest’s
subsampling that give researchers a way of quantifying the uncertainty
of the model’s output (Scornet & Hooker, 2026).

[Impact of an improved random forest-based financial management model on
the effectiveness of corporate sustainability decisions -
ScienceDirect](https://www.sciencedirect.com/science/article/pii/S2772941924000310)

This article gives an example of how they used a random forest model to
help with sustainability decision making in business and to use it as a
better early warning sign for a potential financial crisis. They chose
to use the random forest model due to its low complexity, resistance to
noise data, and good classification performance while also not
increasing the computational effort to much. Then they decided how to
prune the decision trees to be able to reduce the complexity of the
model. They also took their data and normalized it creating Z scores for
each data point to make sure each variable would be treated correctly.
They were able to create the model to have 98.03% accuracy for
predicting financial crisis for companies.

The need for real-time and practical fraud detection systems is
challenging due to the volume of credit card transactions processed.
Credit fraud, and fraud in general, can have negative impacts that are
not limited to financial losses but also include reputational damage
from negative consumer experience.

A recent study (Borketey, 2024) compared several machine learning models
including but not limited to Logistic Regression, Linear Discriminant
Analysis, K-nearest Neighbor, Classification and Regression Trees and
others to try to find the best model to use for real-time fraud
detection. Borketey used a data set with significant imbalance (0.17%
fraudulent transactions). The models were evaluated by using AUC, PRAUC,
F1, KS, Recall and Precision. The data set used was not a full
representation of global credit card fraud since it was specific to
European card issuers. A major limitation is that credit card fraud is
always evolving and this may result in models becoming less effective
over time. The study found the Random Forest model was the best
performing with 71% overall fraud detection.

With the increasing e-commerce transactions, it is important for
financial institutions to develop robust fraud detection and with Random
Forests being a main tool for real-time fraud detection it is important
to understand its strengths and weaknesses. Xuan et al. (2018) compared
two random forest variants Random-Tree-Based Random Forest (Random
Forest I) and Cart-Based Random Forest (Random Forest II).

This experiment looked at how the two models performed, the ratio of
fraudulent to legal transactions, and their performance on a large
dataset. The experiment used a data set from a Chinese e-commerce
company that was imbalanced and had 0.27% fraudulent transactions. Due
to the data imbalance an under-sampling method was used with the
Cart-Based Random Forest to test the ratio of legal and fraudulent
transactions. The test found that the 5:1 ratio provided the best
F-Measure for finding true fraudulent transactions. When this model was
tested on a large data set (4.7 million legal and 38,000 fraudulent
transactions) to simulate real world scenarios the model had a 33%
precision rate. In conclusion, the authors found that RF is good for
smaller data sets; however, further research will be required for real
world application.

Data-set imbalance can be a real concern when it comes to developing
models to detect credit card fraud since these transactions are very
small population for processed transactions. Galar et al. (2012) ran a
series of experiments to provide a taxonomy and running comparisons to
understand how different models handle class imbalance. They used AUC,
which provides a single measurement to evaluate the performace of a
model, compared to accuracy, because it does not distinguish between the
different classifiers. They used a single tree algorithim (C4.5) as the
base classifier. The study reviewed 22 ensemble variants across four
ensemble groups.

The authors  found that the most complex approaches did not always have
the best performance. The most robust methods were SMOTEBagging,
RUSBoost, and UnderBagging. They also took into account the complexity
of the methods and concluded that RUSBoost provided good performance
despite being the least complex. While all ensembles were tested using
the C4.5 classifier it is unclear if these results would remain if
tested against a different base classifier like a Random Forest.

Manual review of fraudulent credit card transaction is arduous and
impractical in today’s financial climate with e-commerce and
cross-border payments. Fraudsters’ tactics are constantly changing, and
this increases the difficulty of identifying and classifying fraudulent
transactions. Fraudulent transactions are a small subset of all
processed transactions resulting in a major data imbalance (Aburbeian &
Ashqar, 2023). The authors used a modified Random Forest that included
synthetic minority oversampling technique (SMOTE) to train the model.
The authors provide the dataset as 94,682 transactions with 87,562
nonfraud (97.7%) and 2,052 (2.3%) fraudulent, causing an inconsistency
in sample calculations. Through testing they found that accuracy is not
the best measure of the model and used a confusion matrix, precision,
recall, F1 score, and Receiver Operating Characteristic (ROC). The
random forest classifier accurately classified 98% of the transactions.
They found that using SMOTE and hyperparameters provided the more
accurate model for addressing the data imbalance with fraudulent
transactions. This experiment relied heavily on synthetic data so
performance might not directly translate to real-world application.

## Methods

- Detail the models or algorithms used.

- Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}$$ $W_n(x)$ *is the sum
of weights that belongs to all real numbers. Weights are positive
numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i$$

## Analysis and Results

### Data Exploration and Visualization

- Describe your data sources and collection process.

- Present initial findings and insights through visualizations.

- Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

- Explain your data preprocessing and cleaning steps.

- Present your key findings in a clear and concise manner.

- Use visuals to support your claims.

- **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

- Summarize your key findings.

- Discuss the implications of your results.

## References

Salman, I. A., Kalakech, A., & Steiti, A. (2024). Random Forest
Algorithm Overview. *Babulonian Journal of Machine Learning*, 69-79.

Scornet, E., & Hooker, G. (2026). Theory of Random Forests. *Annual
Reivew of Statistics and Its Application*, 99-121.
